---
title: "Assignment 10: Data Scraping"
author: "Sarah Slay"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 Load Libraries
library(tidyverse)
library(lubridate)
library(here)
library(rvest) 
# check working directory
here()
```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2024 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 Set the scraping website for the 2024 data
webpage <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024')

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
Sys.name <- webpage %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()

Pwsid <- webpage %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()

Ownership <- webpage %>%
  html_nodes('div+ table tr:nth-child(2) td:nth-child(4)') %>%
  html_text()


Water.sources <- webpage %>%
  html_nodes('tr:nth-child(2) th~ td+ td:nth-child(6) , tr:nth-child(3) th~ td+ td:nth-child(6) , tr:nth-child(4) th~ td+ td:nth-child(6) , tr:nth-child(3) th~ td:nth-child(9) , tr:nth-child(2) td:nth-child(9) , tr:nth-child(2) th~ td:nth-child(3), tr:nth-child(3) th~ td:nth-child(3), tr:nth-child(4) th~ td:nth-child(3), tr:nth-child(4) th~ td:nth-child(9) , tr:nth-child(5) th~ td:nth-child(3), tr:nth-child(5) th~ td:nth-child(9), tr:nth-child(5) th~ td+ td:nth-child(6)') %>%
  html_text()
Water.sources

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... 

5. Create a line plot of the maximum daily withdrawals across the months for 2024, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#4 Create a dataframe for the MGD and ID data
Month.num <- c(1,5,9,2,6,10,3,7,11,4,8,12)
Year <- rep(2024)
MGD.data <- data.frame("Month" = month.abb[Month.num],
                       "Date" = my(paste(Month.num, "-" , Year)),
                        "Water System" = Sys.name,
                        "PWsID" = Pwsid,
                        "Ownership" = Ownership,
                        "Max_Daily_Use_mgd" = as.numeric(Water.sources))

#5 Plot the MGD for each month in 2024
Max_Use.plot <- ggplot(MGD.data,aes(x=Month.num,y=Max_Daily_Use_mgd)) + 
  geom_line()+
  scale_x_continuous(breaks = seq(1, 12, by = 1)) +
  labs(title = paste("2024 Durham Water Use by Month"),
       y="Max Daily Withdrawal (mgd)",
       x="Month")
Max_Use.plot
```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function with two inputs - "PWSID" and "year" - that:
  - Creates a URL pointing to the LWSP for that PWSID for the given year
  - Creates a website object and scrapes the data from that object (just as you did above)
  - Constructs a dataframe from the scraped data, mostly as you did above, but includes the PWSID and year provided as function inputs in the dataframe. 
  - Returns the dataframe as the function's output


```{r construct.a.scraping.function}
#6. Create the scraping function
scrape.web <- function(the_pwsid, the_year) {
  #Construct the scraping web address
  base_url <- "https://www.ncwater.org/WUDC/app/LWSP/report.php"

  scrape_url <- paste0(
    base_url,
    "?pwsid=", the_pwsid,
    "&year=", the_year
  )
  
  #Retrieve the website contents 
  Website <- read_html(scrape_url)

  #set element address variables 
  system_element <- ("div+ table tr:nth-child(1) td:nth-child(2)")
  mdw_element <- c('tr:nth-child(2) th~ td+ td:nth-child(6) , tr:nth-child(3) th~ td+ td:nth-child(6) , tr:nth-child(4) th~ td+ td:nth-child(6) , tr:nth-child(3) th~ td:nth-child(9) , tr:nth-child(2) td:nth-child(9) , tr:nth-child(2) th~ td:nth-child(3), tr:nth-child(3) th~ td:nth-child(3), tr:nth-child(4) th~ td:nth-child(3), tr:nth-child(4) th~ td:nth-child(9) , tr:nth-child(5) th~ td:nth-child(3), tr:nth-child(5) th~ td:nth-child(9), tr:nth-child(5) th~ td+ td:nth-child(6)')
  
  #scrape the elements
  the_mdw <- Website %>% html_nodes(mdw_element) %>% html_text()
  the_system <- Website %>% html_nodes(system_element) %>% html_text()

  
  #Construct a dataframe from the scraped data
  df_function <- data.frame("Month" = month.abb[Month.num],
                            "Water System" = the_system) %>% 
  mutate(PWSID = the_pwsid,
         Year = the_year,
         MDW = as.numeric(the_mdw),
         Date = my(paste(month.abb,"-",Year)))
  df_function
}
                        
```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2020. Then show the structure of the dataframe with either the `str()` or the `glimpse()` function. 

```{r fetch.and.plot.Durham.2020.data}
#7 Use function to plot values from 2020
pan.data <- scrape.web ('03-32-010', 2020)

pan.plot <- ggplot(pan.data,aes(x=Month.num,y=MDW)) + 
  geom_line()+
  scale_x_continuous(breaks = seq(1, 12, by = 1)) +
  labs(title = paste("2020 Durham Water Use by Month"),
       y="Max Daily Withdrawal (mgd)",
       x="Month")
pan.plot
```

8. Use the function above to extract data for Cary (PWSID = '03-92-020') in 2020. Combine this data with the Durham data collected above and create a plot that compares Cary's to Durham's water withdrawals.

```{r fetch.and.plot.Cary.2020.data}
#8 Extract data for Cary by calling function
Cary.data <- scrape.web('03-92-020', 2020)

#combine with durham data
combined.data <- full_join(pan.data, Cary.data)

#plot water withdrawals for comparison
combined.plot <- ggplot() + 
  geom_line(data = pan.data, aes(x = Month.num, y = MDW, color = 'Durham')) +
  geom_line(data = Cary.data, aes(x = Month.num, y = MDW, color = 'Cary')) +
  scale_color_manual(values = c("Durham" = "black",
                                "Cary" = "red")) +
  scale_x_continuous(breaks = 1:12) +
  theme(legend.title = element_text(size = 12)) +
  labs(title = paste("2020 Water Use by Month"),
       y="Max Daily Withdrawal (mgd)",
       x="Month")
combined.plot

```


9. Use the code & function you created above to plot Cary's max daily withdrawal by months for the years 2018 thru 2023.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bind_rows()` to combine the dataframes into a single one, and use that to construct your plot.

```{r fetch.and.plot.Cary.multiyear.data }
#9 Call Cary data from 2018 thru 2023 using function
Year_range <- rep(2018:2023)
Cary.loc <- "03-92-020"

#iterate through the dataframe years and combine into one dataframe
df_iterate <- map(Year_range,scrape.web,the_pwsid = Cary.loc)
df_total <- bind_rows(df_iterate)

#plot the combined data
Combo.plot <- ggplot(df_total,aes(x=Year,y=MDW)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(
    title = paste(
      "Cary Water usage:",Year_range[1],"-",Year_range[length(Year_range)]
      ),
    y="Max Daily Withdrawal (mgd)",
    x="Month"
    )
Combo.plot
```

Question: Just by looking at the plot (i.e. not running statistics), does Cary have a trend in water usage over time?
> Answer: Cary's water usage has increaed slightly from 2018 to 2023, which correlates with population growth. There is a noticable drop in water use in 2020 likely due to the pandemic. The largest increase happened between 2020 and 2021. 
>
