---
title: "Assignment 2: Coding Basics"
author: "Sarah Slay"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons/labs in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Canvas.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 55, increasing by fives. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. Creates object named OneFive to list numbers 1 through 55 by fives using the sequence function.
    OneFive <- seq(1, 55, by=5)
    OneFive

#2. Gives mean and median of object OneFive using mean and median functions.
    mean1 <- mean(OneFive)  
    median1 <- median(OneFive)

#3. Determine whether mean is greater than median using conditional statement
    mean1 > median1
    
```

## Basics, Part 2

5.  Create three vectors, each with four components, consisting of (a) student names, (b) test scores, and (c) whether they are on scholarship or not (TRUE or FALSE).

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}
    studentname <- c("Simon Beery", "Jen Watkins", "Peter Philips", "Ben Smith") #Vector type: String
    testscores <- c(89, 94, 92, 91) #Vector type: Numeric
    scholarship <- c(TRUE, TRUE, TRUE, FALSE) #Vector type: Boolean
    
    StudentInfo <- data.frame(studentname,testscores,scholarship) 
    names(StudentInfo) <- c("Student Name","Test Score","Scholarship")
    StudentInfo
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: Matrices combine multiple vectors of equal length and mode. Data frames combine multiple vectors into tables that can have different modes.

10. Create a function with one input. In this function, use `if`...`else` to evaluate the value of the input: if it is greater than 50, print the word "Pass"; otherwise print the word "Fail". 

11. Create a second function that does the exact same thing as the previous one but uses `ifelse()` instead if `if`...`else `. 

12. Run both functions using the value 52.5 as the input

13. Run both functions using the **vector** of student test scores you created as the input. (Only one will work properly...)

```{r basics 3}
#10. Create a function using if...else
  GreaterThan50 <- function(x) {
  if(x > 50) {
    print("Pass")
  } 
  else {
    print("Fail")
  }
}


#11. Create a function using ifelse()
GreaterThan50IE <- function(x){
  ifelse(x>50, print("Pass"), print("Fail"))
}

#12a. Run the first function with the value 52.5
GreaterThan50(52.5)

#12b. Run the second function with the value 52.5
GreaterThan50IE(52.5)

#13a. Run the first function with the vector of test scores
#GreaterThan50(testscores) Commented out because code error

#13b. Run the second function with the vector of test scores
GreaterThan50IE(testscores)
```

14. QUESTION: Which option of `if`...`else` vs. `ifelse` worked? Why? (Hint: search the web for "R vectorization")

> Answer: The ifelse option worked on the test scores vector because it applies the test condition to each element in a vector sequentially. Then, ifelse returns a true or false response for each element of the vector. In contrast, If... else can only handle one value at a time. 


**NOTE** Before knitting, you'll need to comment out the call to the function in Q13 that does not work. (A document can't knit if the code it contains causes an error!)